Type-Safe Modular Parsing

### 2016-12-08 paper/sections/Overview.tex
First comments on Section 2

\subsection{The Choice of Parsing Techniques}\label{subsec:overview-parsing}

Parsing techniques have been heavily studied over the years, especially for context-free grammars and its subsets. However, the choice of parsing techniques is not a minor issue in a modular setting. To achieve type-safe modularity, the parsers must be able to be extended and composed, and different parsing techniques have different suitability for that.

Although every parsing algorithm can be implemented by hand in principle, some of them are not friendly for that. A \textit{parser generator} is a tool to help users generate parsers automatically or semi-automatically from a given grammar. There is no restriction on the algorithm, while most of them adopts table-based LL and LR parsing algorithms.

\huang{a paragraph here, refer to related work section, discuss previous works of modular parsing using parser generators and its drawbacks}

Comparing with the parser generators, a \textit{parser combinator} takes several parsers and produce a new parser as its output. It is popular in functional programming, where the parsers are represented by functions and parser combinators are higher-order functions accepts them.

Parser combinators are very suitable for our purpose, because of two reasons. Firstly, they are naturally modular. The manner of using them is to write small parsers and use combinators to composed them together. The construction procedure is explicit and fully controlled by the programmer. Secondly, each parser combinator is represented by a piece of code, and also are the parsers it takes. Thus in a statically typed programming language they can be type-checked during compile time.

Parser combinators adopt top-down, recursive descent parsing, which uses backtracking to search through the possible branches. However, simple backtracking parser combinators such as the famous library Parsec in Haskell, have some drawbacks. One is that they cannot support left-recursive grammars. The common solution is to rewrite a left-recursive grammar into an equivalent one, so called \textit{left-recursion elimination}. This solution requires information of the whole grammar, so that it cannot be applied in a modular setting. Because the grammar may be extended in the future, essentially we only have parts of the grammar. Another drawback is performance. Take Parsec as an example, its choice combinator only tries the second alternative if the first fails without any token consumption. If two alternatives have a same non-empty prefix, an auxiliary function \lstinline{try} must be added to backtrack. However, if we want to extend our parsers later, we should always consider the worst case in which all alternatives share common prefixes. Then we need to add \lstinline{try} for all the branches, which results worst case exponential time complexity.

Beyond the simple backtracking one, several advanced parser combinators were studied to resolve those issues. One of them is Packrat parsers, which is our choice. The details of Packrat parser are beyond the scope of this paper. In brief, Packrat parsers use memorization to record the result of applying each parser at each position of the input, so that repeated computation is eliminated, and it supports left-recursive grammars. All of these properties are very suitable for modularity, thus we use Packrat parsers as the underlying parsing technique. In section \ref{sec:packratparsers}, we will talk about the Packrat parsing library implemented in Scala, which we use.

@@

\subsection{Modular Parsers, the First Look}\label{subsec:overview-firstlook}

In this section we start demonstrating how to write modular parsers. Suppose we want to parse a simple language of variables and applications.

\begin{lstlisting}[language=PlainCode]
Expr := Var String | App Expr Expr
\end{lstlisting}

The abstract syntax can be defined by Scala's case classes, and then we write parsers for the two cases \lstinline{Var} and \lstinline{App}.

@@

val pApp: (=> PackratParser[Expr]) => PackratParser[Expr] =
  e => e ~ e ^^ { case e1 ~ e2 => App(e1, e2) }
\end{lstlisting}

Some structures are imported from the Scala Packrat parsing library, such as the combinator \lstinline{^^}. They will be discussed in later sections, so that readers could only focus on the type of parsers at the first reading.

\lstinline{pVar} for parsing variables is straight forward. It parses a single identifier, and its type \lstinline{PackratParser[Expr]} indicates the result is an \lstinline{Expr}.

The type of \lstinline{pApp} is different. It takes a parser of \lstinline{Expr} as an argument, and use it to parse two consecutive expressions as an application. This argument \lstinline{e} represents the parser of \emph{the whole expression}. In this particular case, it is in fact the combination of \lstinline{pVar} and \lstinline{pApp}. We can replace it by \lstinline{pVar | pApp} and the code still works. However, using an abstract argument provides extensibility to the parser. Because once the abstract syntax is extended later, the parser of the whole expression will change correspondingly. Then we have to edit existing code if it is hard-coded. The parameterization technique used in \lstinline{pApp} is called \textit{open recursion}, which is discussed in section \ref{sec:openandparsing}.

We can use a trait to glue them together, and define \lstinline{pVarApp} as a combination of the two parsers. A type synonym \lstinline{Fix} is used to shorten the types.

@@

  val pVar: PackratParser[Expr] =
    ident ^^ Var
  val pApp: Fix[PackratParser[Expr]] =
    e => e ~ e ^^ { case e1 ~ e2 => App(e1, e2) }
  val pVarApp: Fix[PackratParser[Expr]] =
    e => pApp(e) | pVar | "(" ~> e <~ ")"
}
\end{lstlisting}

Finally, we are able to parse \lstinline{Expr} frome strings as shown in the code below. The \lstinline{fix} function here is just the standard \textit{fixpoint}, it is explained in section \ref{subsec:openrecursion}.

@@

\subsection{Extend and Reuse Parsers}\label{subsec:overview-extend}

We have done our modular parser, and it behaves just in a normal way. To demonstrate the advantages of writing modular parsers, we extend the language to add lambda abstractions as a new case of expressions. The language then becomes the famous \textit{untyped lambda calculus}.

\begin{lstlisting}[language=PlainCode]
Expr := Var String | App Expr Expr | Lam String Expr
\end{lstlisting}

@@

\subsection{Object Algebras and More}\label{subsec:overview-oa}

>>>>

\subsection{Chosing the Parsing Technique}\label{subsec:overview-parsing}

%A technique for type-safe modular parsing should the following 3 
%features: \emph{modular type-checking}; \emph{separate compilation}; 
%low performance overhead

Although there are many parsing techniques, not all of them are
suitable for type-safe modular parsing. In particular there are many 
techniques which fail to provide modular type-checking and separate 
compilation. Moreover, even if modular type-checking and separate 
compilation are supported, efficiency is another
concern. A parsing technique should have low overhead when applied 
in a modular setting. In the remaider of this section, we eliminate 
various techniques that fail to satisfy our requirements, and argue
that that Packrat parsing~\cite{} is a suitable candidate for
type-safe modular parsing.

\paragraph{Parser Generators} The most widely use tools for parsing 
are parser generators~\cite{}. Parser generators help users generate parsers automatically or
semi-automatically from a given grammar. There is no restriction on
the algorithm, while most of them adopts table-based LL~\cite{} and LR parsing
algorithms~\cite{}. 
Although efficient, the main drawback of parser generators is that they do not support 
modular type-checking and separate compilation.

\huang{a paragraph here, refer to related work section, discuss
  previous works of modular parsing using parser generators and its
  drawbacks}\bruno{Please finish this.}

\paragraph{Parser Combinators}
Comparing with the parser generators, a \textit{parser combinator}~\cite{}
takes several parsers and produce a new parser as its output. Parser combinators are
popular in functional programming, where the parsers are represented
by functions and parser combinators are higher-order functions accepts
them.

At a first look, parser combinators are very suitable for our purpose, because of two
reasons. Firstly, they are naturally modular. The manner of using them
is to write small parsers and use combinators to composed them
together. The construction procedure is explicit and fully controlled
by the programmer. Secondly, each parser combinator is represented by
a piece of code, and also are the parsers it takes. Thus in a
statically typed programming language they can be statically
type-checked.

Unfortunatelly many parser combinators have important limitations for
modular parsing. In particular several parser combinators, 
including the famous library Parsec~\cite{} library, require
programmers to manually do\textit{left-recursion elimination}, and
require significant amounts \textit{backtracking}. Both of these are
problematic in a modular setting.

\paragraph{Left-Recursion Elimination} The main problem with
left-recursion elimination is that it is a \emph{global}
transformation on a grammar. Given that the full grammar is known,
then it is possible to remove all left-recursive cases.
\bruno{Illustrate with a concrete example: show a simple
  left-recursive grammar; then do left-recursion elimination. Then add
a new case into the original left-recursive grammar; and show that 
a very different grammar is obtained after left-recursion elimination.}

However, when doing modular parsing, the full grammar is not known!

Parser combinators adopt top-down, recursive descent parsing, which
uses backtracking to search through the possible branches. However,
simple backtracking parser combinators such as the famous library
Parsec~\cite{} in Haskell, have some drawbacks. One is that they cannot
support left-recursive grammars. The common solution is to rewrite a
left-recursive grammar into an equivalent one, so called
\textit{left-recursion elimination}. This solution requires
information of the whole grammar, so that it cannot be applied in a
modular setting. Because the grammar may be extended in the future,
essentially we only have parts of the grammar.

\paragraph{Backtracking} The need for backtracking is also problematic
in a modular setting. Take Parsec as an example, its choice
combinator only tries the second alternative if the first fails
without any token consumption. If two alternatives have a same
non-empty prefix, an auxiliary function \lstinline{try} must be added
to backtrack. However, if we want to extend our parsers later, we
should always consider the worst case in which all alternatives share
common prefixes. Then we need to add \lstinline{try} for all the
branches, which results worst case exponential time complexity.
\bruno{you should again illustrate your point here with an
  example. The example should illustrate that the addition of a new
  case may require introducing backtracking. }

\paragraph{Packrat Parsing} 
Fortunatelly more advanced parsing techniques such as Packrat
parsing~\cite{}, address the limitations of simple parser combinators 
such as Parsec. Packrat parsers use
memorization to record the result of applying each parser at each
position of the input, so that repeated computation is eliminated, and
they supports left-recursive grammars directly. All of these properties are very
suitable for modularity, thus we use Packrat parsers as the underlying
parsing technique. In section \ref{sec:packratparsers}, we will give
an overview of the Packrat parsing library implemented in Scala, 
which we use.

@@

\subsection{The Parsing Expression Problem}
\bruno{You are not motivating the problem! You are going straight to
  the solution without pointing out what the problem is first. What
  you need to do is: First show what happens with conventional
  parsers: at some point, if you add extensions the recursive calls
  will be wrong. Then you show (in the next section) the solution: 
 use delegation/open recursion.}

\bruno{I think we need to set up a challenge here, similar to the challange of
  the expression problem. The challange should be like. Build a parser
for a simple expression language, then extend the language with both 
a new language construct and a new operation. This section will show 
how we can do that using traditional parsers, but without modular
type-safety and separate compilation. The remaining sections will show
that the techniques introduced by us, enable us to solve those two challenges.}

Typical parsers defined parser combinators follow the grammar
structure, and build a corresponding AST. If the grammar is changed to
add a new alternative, the corresponding parser has to be changed as
well. Thus the traditional style of developing parsers is not
extensible, and it is at odds with object-oriented programming ASTs,
which support the easy addition of new alternative classes. We
illustrate this problem with a simple expression language and extension.

Suppose we want to parse a simple language of variables and
applications.

\begin{lstlisting}[language=PlainCode]
Expr := Var String | App Expr Expr
\end{lstlisting}
\bruno{if you intend to write a grammar here, then please use grammar
  (BNF) notation, and not some notation invented by you!}

\paragraph{Extensing the Parser}

\paragraph{Problem: Hardcoded Recursive Calls}

\subsection{Delegation-Based Parsing}\label{subsec:overview-firstlook}

The problem of hardcoded recursive calls can be solved using
delegation, which can be encoded with open recursion. The idea is
simple: instead of making direct recursive calls, an additional
argument, that abstracts over the recursive call, is passed to the 
parsers. This technique enables the developement of parsers that 
support OO-style extensibility of ASTs.

\paragraph{Open Parsing}
The example in Section \ref{} can be rewritten as follows:

@@

val pApp: (=> PackratParser[Expr]) => PackratParser[Expr] =
  e => e ~ e ^^ { case e1 ~ e2 => App(e1, e2) }
\end{lstlisting}
\bruno{You are copying and pasting code into the paper. Don't do this:
code will change alot (various edits), and it is all too easy to get
inconsistent code snippets. Instead use the small Ruby tool that we 
have in the paper directory to import source code directly from the
code repository. This will ensure the consistency of the code
throughout the paper writing. Please ask Haoyuan about the tool; he
knows how to use it!}
\bruno{Please polish up code. You probably want to consistently use 
the variable name ``self'' whenever you want to refer to the whole
parser (instead of ``e'' that you use at the moment). Also you should 
use ``Open'' (which you currently call ``Fix'') instead of using the
longwinded ``(=> PackratParser[Expr]) => PackratParser[Expr]''} 
\bruno{Also, for consistency, why isn't ``pVar'' open? I.e. using a
  self reference as well? I know that it is not recursive, but giving
  it a different treatment makes it more confusing for users.}
\bruno{You should use a more OO approach in the example. 
Rather than case classes show an Expr trait with pretty printing.
This will enable you to talk about the extensibility and object
algebras better later on.
}

The abstract syntax is defined by Scala's case classes, and 
we write parsers for the two cases \lstinline{Var} and \lstinline{App}.
The parser \lstinline{pVar}, for parsing variables, is straightforward. It parses a
single identifier, and its type \lstinline{PackratParser[Expr]}
indicates the result is an \lstinline{Expr}.
The parser \lstinline{pApp}, for parsing applications is more
interesting. \lstinline{pApp} takes a parser of
\lstinline{Expr} as an argument, and uses it to parse two consecutive
expressions as an application. This argument \lstinline{e} represents
the parser of \emph{the whole expression}. In this particular case, it
is in fact the combination of \lstinline{pVar} and
\lstinline{pApp}. We can replace it by \lstinline{pVar | pApp} and the
code still works. However, using an abstract argument provides
extensibility to the parser. Because once the abstract syntax is
extended later, the parser of the whole expression will change
correspondingly. Then we have to edit existing code if it is
hard-coded. The parameterization technique used in \lstinline{pApp} is
called \textit{open recursion}, which is discussed in section
\ref{sec:openandparsing}.

\paragraph{Parsing Components}\bruno{consider removing and/or moving
  to later parts of the paper}
We can use a trait to glue them together, and define
\lstinline{pVarApp} as a combination of the two parsers. A type
synonym \lstinline{Fix} is used to shorten the types.

@@

  val pVar: PackratParser[Expr] =
    ident ^^ Var
  val pApp: Fix[PackratParser[Expr]] =
    e => e ~ e ^^ { case e1 ~ e2 => App(e1, e2) }
  val pVarApp: Fix[PackratParser[Expr]] =
    e => pApp(e) | pVar | "(" ~> e <~ ")"
}
\end{lstlisting}
\bruno{Fix is not the right name. A better name is Open (for open
  recursion).}
\bruno{I'm not sure that showing the code for language components is
  essential at this point. This can be shown latter in the paper.}

\paragraph{Client Code}
Finally, we are able to parse \lstinline{Expr} frome strings as shown in the code below. The \lstinline{fix} function here is just the standard \textit{fixpoint}, it is explained in section \ref{subsec:openrecursion}.

@@

%\subsection{Extend and Reuse Parsers}\label{subsec:overview-extend}

\paragraph{Extensibility} To demonstrate the advantages of writing 
parsers with open recursion, we extend the language to add lambda 
abstractions as a new case of expressions. The language then 
becomes the famous \textit{untyped lambda calculus}.

\begin{lstlisting}[language=PlainCode]
Expr := Var String | App Expr Expr | Lam String Expr
\end{lstlisting}
\bruno{same comment as before: if this is supposed to be a grammar,
  then use a grammar notation}

@@

\subsection{Object Algebras for full Extensibility}\label{subsec:overview-oa}

Although delegation enables OO extensibility, the use of an OO class
hierarchy makes the addition of new operations over the AST
problematic. Object Algebras~\cite{} enable us to solve this problem, and
offer high flexibility in the choice of operations to be performed
over the AST.
